# Ordered Factors

```{r}
#| label: libraries
#| message: false
#| echo: false
#| warning: false

library(tidyverse)
library(primer.data)
library(tidymodels)
library(broom)
library(marginaleffects)
```

<!-- IS: add more intro than just the data? -->

The data that we will use was downloaded from Opportunity Insights. Based at Harvard University, Opportunity Insights focuses on using big data to understand and improve economic mobility. They have many downloadable data sets, check it out for yourself [here!](https://opportunityinsights.org/data/)

Data science is ultimately a moral act, so we will use the four [Cardinal Virtues](https://en.wikipedia.org/wiki/Cardinal_virtues) --- Wisdom, Justice, Courage and Temperance --- to guide our approach.  

## Wisdom

Wisdom begins with examining the data at hand, that can ultimately lead to our question. We must create a Preceptor Table to answer the question of what data would we, ideally, require to answer our question. It is then that we can explore the data that we actually have.

### EDA

<!-- IS: should this come after the Preceptor Table? -->

<!-- needs more exploration of data -->

Consider the `colleges` data set. `colleges` contains data from over 900 colleges and universities in the United States. These draw primarily on data from the Department of Education’s (DOE) [IPEDS](https://nces.ed.gov/ipeds/) database in 2013 and the College Scorecard. 

```{r}
#| code-fold: false
#| warning: false
library(tidyverse)
library(primer.data)

glimpse(colleges)
```

Given colleges are the broad topic that we are looking for, we will use `grad_rate` as our outcome variable.

```{r}
#| code-fold: false
colleges |> 
  select(name, grad_rate)
```

When considering a predictive model, `tuition` is an important link to `grad_rate`, as people who pay more for a college may be more willing to spend their time there.

```{r}
#| code-fold: false
colleges |> 
  select(name, grad_rate, tuition)
```

<!-- IS: this transition is awkward, not sure how to introduce our overall question (maybe at the end with the causal model?) -->

From here we can consider our central predictive model question:

> *What effect does the tuition of a college have on its graduation rate?*

When considering a causal model, we must think about variables that we can easily manipulate to get our two potential outcomes. In this case, we are observing what variables would impact tuition if changed. This additionally means that we are changing our outcome variable to tuition. We could consider observing tuition if the school is in the South rather than in the North, using the variable `region`.

```{r}
#| code-fold: false
colleges |> 
  filter(region %in% c("North", "South")) |> 
  select(name, tuition, region)
```

Of course, then we might account for the other regions, and our model would be cluttered with the number of treatment variables. Instead, we consider the `public` variable, observing whether or not the school is private or public and how that impacts the overall tuition.

```{r}
#| code-fold: false
colleges |> 
  select(name, tuition, public)
```

This is a much easier variable to manipulate, as we can easily set a school to either be private or public. 

From here we can consider our central causal model question:

> *How does a college's public or private status impact tuition?*

Now that we have explored the data and know what we are looking for in our questions, we can create our Preceptor Tables.

### Preceptor Table

<!-- IS: when observing predictive vs causal, should we have seperate sections? -->

First, we will observe and create the Preceptor Table for our predictive model.

Which rows and columns of data do you need such that, if you had them all, the calculation of the quantity of interest would be trivial? The steps we usually take to construct the Preceptor Table include:

*Units*: Each individual college in the United States.

*Outcome*: The `grad_rate` for each individual college.

*Treatment*: Since we are focusing on our predictive model, there are no treatment variables.

*Covariates*: Obviously we must include `tuition`, however variables like `selectivity` or `tier` (prestige) are useful in helping us figure out more about the college.

*Moment in Time*: Now. Although, recall that a Preceptor Table can never really refer to an exact instant in time since nothing is instantaneous in this fallen world.

Predictive models care nothing about causation. Causal models are often also concerned with prediction, if only as a means of measuring the quality of the model. Here, we are looking at *prediction*. 

We can now build our Preceptor Table and visualize it. Since we are predicting graduation rate for every college in America, we must include a row for each college. As of 2025 there are approximately 5,999 colleges in the U.S., so we must have 5,999 rows.

Here is our Preceptor Table for our predictive model:

```{r}
library(gt)

data <- tibble(
  ID = c("1", "2", "...", "10", "11", "...", "N"),
  graduation_rate = c(0.92, 0.55, "...", 0.42, 0.78, "...", 0.87),
  tuition = c(4.5, 1.3, "...", 2.9, 2.2, "...", 3.7)
)
data |>
  gt() |>
  tab_header(title = "Preceptor Table") |>
  cols_label(
    ID = md("ID"),
    graduation_rate = md("Graduation Rate"),
    tuition = md("Tuition (tens of thousands)")
  ) |>
  tab_style(
    cell_borders(sides = "right"),
    location = cells_body(columns = c(ID))
  ) |>
  tab_style(
    style = cell_text(align = "left", v_align = "middle", size = "large"), 
    locations = cells_column_labels(columns = c(ID))
  ) |>
  cols_align(align = "center", columns = everything()) |>
  cols_align(align = "left", columns = c(ID)) |>
  fmt_markdown(columns = everything()) |>
  tab_spanner(label = "Outcome", columns = c(graduation_rate)) |>
  tab_spanner(label = "Covariate", columns = c(tuition))

```

This table would extend all the way for all the colleges. We don't know the number of rows, so we use `N`. If we had this table, all of our questions could be answered with simple math. No inference is necessary if we have a Preceptor Table.